---
title: "Crop Statistics 1900-2017 dataset analasys"
author: "Jetze Luyten, Axel Van Gestel, David Silva Troya"
output:
  pdf_document: default
---

## Setup the enviroment
The first thing to setup for the analysts is the environment with the required packages and settings.

### Install required packages and load required libraries
During this analysis we used the `tidyverse` package for reading, cleaning and plotting the data and the `ggcorrplot` package to visualize the correlation matrix into a heat map.
```{r message=FALSE}
# install.packages("tidyverse")
# install.packages("ggcorrplot")

library(tidyverse) # Contains all tidyverse packages (ggplot2, dplyr, ...)
library(ggcorrplot) # Used for generating correlation heatmaps (uses ggplot2)
```


### Setup enviroment settings
In the following code block we set the language R uses for it's messages to English, clear all the global variables so that we always start with a clean slate and setup ggplot to center the plot titles by default.
```{r}
Sys.setenv(LANG = "en") # Set language to English
rm(list = ls()) # Clears the Global Env
theme_update(plot.title = element_text(hjust = 0.5)) # Center all plot titles
```
