---
title: "Crop Statistics 1900-2017 dataset analasys"
author: "Jetze Luyten, Axel Van Gestel, David Silva Troya"
output:
  pdf_document: default
---

## Setup the enviroment
The first thing to setup for the analysts is the environment with the required packages and settings.

### Install required packages and load required libraries
During this analysis we used the `tidyverse` package for reading, cleaning and plotting the data and the `ggcorrplot` package to visualize the correlation matrix into a heat map.
```{r message=FALSE}
# install.packages("tidyverse")
# install.packages("ggcorrplot")
library(tidyverse) # Contains all tidyverse packages (ggplot2, dplyr, ...)
library(readxl) #  Need to load explicitly (not a core tidyverse package)
library(ggcorrplot) # Used for generating correlation heatmaps (uses ggplot2)
library("scales")

```


### Setup enviroment settings
In the following code block we set the language R uses for it's messages to English, clear all the global variables so that we always start with a clean slate and setup ggplot to center the plot titles by default.
```{r}
Sys.setenv(LANG = "en") # Set language to English
rm(list = ls()) # Clears the Global Env
theme_update(plot.title = element_text(hjust = 0.5)) # Center all plot titles
```



## Read and import the data set
### Read the data set (uses readr)
```{r}

column_types <- c(
  "numeric", # ...1
  "numeric", # Harvest_year
  "text",    # admin0
  "text",    # admin1
  "text",    # crop
  "numeric", # hectares (ha)
  "numeric", # production (tonnes)
  "numeric", # year
  "numeric", # yield(tonnes/ha)
  "text",    # admin2
  "text"     # notes
)
crops <- read_xlsx(
  path = "./crops/food-twentieth-century-crop-statistics-1900-2017-xlsx.xlsx",
  sheet = "CropStats",
  col_types = column_types)

```

## Drop not needed columns
```{r}
crops <- select(crops, -c(...1, admin2, notes, Harvest_year))

crops <- crops %>% mutate(crop = factor(crop,
                                        levels = c("wheat", "winter wheat",
                                                   "spring wheat", "maize",
                                                   "cereals"),
                                        ordered = TRUE))

crops <- crops %>% mutate(year = as.integer(year))
```

## Clear not needed variables
```{r}
rm(column_types)
```

# Filtering and cleaning
### Check for the number of NA's in each column
```{r}
sanity_check <- function(my_df) {
  for (j in 1:ncol(my_df)) {
    print(paste(names(my_df[j]), ":", sum(is.na(my_df[, j]))))
  }
}

sanity_check(crops)


```

### View 'crop' tibble
```{r}
crops
```



## Correlation heatmap (uses ggcorrplot)
### Generate a correlation heatmap of the numeric values
```{r message=FALSE, warning=FALSE}
crops_numeric <- select(crops,
                        `hectares (ha)`,
                        `production (tonnes)`,
                        year ,
                        `yield(tonnes/ha)`)

crops_numeric_corr <- cor(crops_numeric, use = "complete.obs") # Use only non NA

ggcorrplot::ggcorrplot(crops_numeric_corr,
                       lab = TRUE, # Show correlation coefficients
                       colors = c("darkturquoise", "white", "salmon"),
                       title = "Correlation between the numeric values")

# Clear not needed variables
rm(crops_numeric, crops_numeric_corr)
```


## Plots and stuff (uses ggplot2)
### Production x years
```{r message=FALSE, warning=FALSE}
ggplot(data = crops, mapping = aes(x = year, y = `production (tonnes)`)) +
  geom_point(alpha = 1/10) +
  stat_smooth(method = "lm", formula = y ~ x + I(x^2), size = 1)+
  ggtitle("Production over the years")
```

### Yield outliers
```{r message=FALSE, warning=FALSE}
ggplot(data = crops, mapping = aes(x = crop, y = `yield(tonnes/ha)`)) +
  geom_boxplot() +
  ggtitle("Yield outliers")
```


### Count of crops
```{r message=FALSE, warning=FALSE}
ggplot(data = crops, mapping = aes(x = crop, y = `yield(tonnes/ha)`)) +
  geom_boxplot() +
  ggtitle("Yield outliers")
```


### ?
```{r message=FALSE, warning=FALSE}
ggplot(data = crops, mapping = aes(x = crop, y = `hectares (ha)`)) +
  geom_boxplot() +
  scale_y_continuous(labels = comma) +
  ggtitle("Hectares outliers")
```











## References
* [Correlation heatmap using ggplot2](http://www.sthda.com/english/wiki/ggcorrplot-visualization-of-a-correlation-matrix-using-ggplot2)





