---
title: "Titanic dataset analasys"
author: "Jetze Luyten, David Silva Troya, Axel Van Gestel"
output:
  word_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

## Setup the enviroment
### Setup settings and install required packages
```{r}
Sys.setenv(LANG = "en") # Set language to English
setwd(getwd()) # Set the working directory to the script directory
rm(list = ls()) # Clears the Global Env

# install.packages("tidyverse")
```


### Load required libraries
```{r message=FALSE}
library(tidyverse) # Contains all tidyverse packages (ggplot2, dplyr, ...)
```



## Read and import the data set
### Read the data set (uses readr)
```{r}
column_types <- cols(
  Survived = col_factor(),
  Pclass = col_factor(include_na = TRUE, ordered = TRUE),
  Sex = col_factor(),
  Embarked = col_factor(include_na = TRUE, ordered = TRUE)
)
train <- read_csv("./kaggle/titanic/train.csv", col_types = column_types)
```

### Rename the factors to be human readable (uses dplyr)
```{r}
train$Survived <- recode_factor(train$Survived,
                                "0" = "No",
                                "1" = "Yes")

train$Pclass <- recode_factor(train$Pclass,
                              "1" = "1st",
                              "2" = "2nd",
                              "3" = "3rd",
                              .default = "Unknown", # NA -> Unknown
                              .ordered = TRUE)

train$Embarked <- recode_factor(train$Embarked,
                                "S" = "Southampton (England)",
                                "C" = "Cherbourg (France)",
                                "Q" = "Queenstown (Ireland)",
                                .default = "Unknown", # NA -> Unknown
                                .ordered = TRUE)
```



## Filtering and cleaning
### Check for the number of NA's in each column
```{r}
sanity_check <- function(my_df) {
  for (j in 1:ncol(my_df)) {
    print(paste(names(my_df[j]), ":", sum(is.na(my_df[, j]))))
  }
}

sanity_check(train)
```

### View 'train' tibble
```{r}
train
```



## Adding useful columns
### Add a total Family size column
```{r}
train <- mutate(train, FamilySize = SibSp + Parch)
```

### Group the cabin label into has cabin and has no cabin
```{r}
train <- mutate(train, CabinGroups = ifelse(is.na(train$Cabin),
                                            "No cabin",
                                            "Cabin"))
```

### Quick sanity check of the 'train' tibble
```{r}
tail(train)
```



## Plots and stuff (uses ggplot2)
### Embarked and Fare prices
```{r message=FALSE, warning=FALSE}
ggplot(data = train, mapping = aes(x = Embarked, y = Fare)) +
  geom_boxplot()
```

### Pclass, Family size and Survived bigger than fare 500
```{r message=FALSE, warning=FALSE}
FareEnough <- filter(train, Fare > 500) # Fare bigger than 500

ggplot(data = FareEnough, mapping = aes(x = Pclass, y = FamilySize)) +
  geom_point(aes(shape=Survived))
```

### Count of family size who paid over 500
```{r message=FALSE, warning=FALSE}
ggplot(data = FareEnough, mapping = aes(x = FamilySize)) +
  geom_histogram()
```

### Male Female survival percentage
```{r message=FALSE, warning=FALSE}
ggplot(data = train, mapping = aes(x = Sex, fill = Survived)) +
  geom_bar(position = "fill")
```

### Pclass survival percentage
```{r message=FALSE, warning=FALSE}
ggplot(data = train, mapping = aes(x = Pclass, fill = Survived)) +
  geom_bar(position = "fill")
```

### Pclass and cabin label percentage
```{r message=FALSE, warning=FALSE}
ggplot(data = train, mapping = aes(x = Pclass, fill = CabinGroups)) +
  geom_bar(position = position_fill(reverse = TRUE)) +
  scale_fill_manual(values = c("darkturquoise",
                               "salmon"))
```

### Male FamilySize survival percentage
```{r message=FALSE, warning=FALSE}
ggplot(data = train, mapping = aes(x = FamilySize, fill = Survived)) +
  geom_bar(position = "fill") +
  facet_wrap(~ Sex) +
  scale_x_continuous(breaks = unique(train$FamilySize))
```
